\textsf{Are large language models superhuman chemists?}

### 2024-10-28 src/tex/appendix.tex
small fix of text

\begin{table}
    \centering
    \small
    \caption{\textbf{Examples for each of the topics considered in the evaluation of the \chembench corpus.} The table shows the percentage of questions in the corpus that belong to each topic, as well as example questions.}

@@

    {\fontsize{8}{9}\selectfont
        \begin{tabularx}{\textwidth}{X}
            \multicolumn{1}{c}{\textbf{Analytical Chemistry} 166 Questions (5.8\%)} \\
            Which of the following analytical methods is most appropriate for performing a survey analysis of a solid sample containing various metals? \\
            A. X-ray fluorescence analysis \\
            B. Differential pulse polarography \\
            C. Flame-atomic absorption spectroscopy \\
            D. Gas chromatography with flame ionization detector \\
            E. Hydride generation atomic absorption spectroscopy \\
            \multicolumn{1}{c}{\textbf{Chemical Preference} 1001 Questions (35\%)} \\
            Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
            A. [START\_SMILES]N\#Cc1ccc(OCCCN2CC3CN\-(CCNS(=O)(=O)c4ccc(F)cc4)CC(C2)O3)cc1\-[END\-\_SMILES] \\
            B. [START\_SMILES]O=C1CC(c2ccc(CC(NS(=O)\-(=O)\-c3cc(Cl)cc(Cl)c3)c3nc4ccccc4[nH]3)cc2)S\-(=O)\-(=O)N1[END\_SMILES] \\
            \multicolumn{1}{c}{\textbf{General Chemistry} 152 Questions (5.3\%)} \\
            Which of the following salts is an acidic salt? \\
            A. \ce{NH4Cl} \\
            B. \ce{Na2CO3} \\
            C. \ce{NaH2PO4} \\
            D. \ce{Zn(OH)Cl} \\
            \multicolumn{1}{c}{\textbf{Inorganic Chemistry} 94 Questions (3.3\%)} \\
            What is the oxidation number of the metal in the compound \ce{[ZrF_{7}]3-}? \\
            \multicolumn{1}{c}{\textbf{Materials Science} 89 Questions (3.1\%)} \\
            For NMR analysis, you need to digest the MOF in a strong acid to remove the linker and leave the metal clusters intact. Why would one choose HF over \ce{HCl} for this purpose? \\
            A. \ce{F-} forms a stable bonds to the metal ions \\
            B. \ce{HF} has a better water solubility than \ce{HCl} \\
            C. \ce{HF} has a higher boiling point than \ce{HCl} \\
            D. \ce{HF} is a weaker acid than \ce{HCl} \\
            \multicolumn{1}{c}{\textbf{Organic Chemistry} 441 Questions (15.5\%)} \\
            What is the reaction mechanism that describes the following reaction (represented using reaction SMILES) [START\_RXNSMILES]CCCl.CO[Na]>>[Na]Cl.CCOC\-[END\_RXNSMILES]? \\
            A. $E_1$ \\
            B. $E_{cb}$ \\
            C. $S_N1$ \\
            D. $S_N2$ \\
            \multicolumn{1}{c}{\textbf{Physical Chemistry} 179 Questions (6.3\%)} \\
            The Born-Oppenheimer (BO) approximation is widely used in computational chemistry, but its accuracy can vary depending on the system. Among the following options, for which system is the Born-Oppenheimer approximation likely to be least applicable? \\
            A. \ce{C60} \\
            B. \ce{CH4} \\
            C. \ce{Fe(CO)5} \\
            D. \ce{H2+} \\
            E. \ce{NaCl} \\
            \multicolumn{1}{c}{\textbf{Technical Chemistry} 42 Questions (1.5\%)} \\
            Which of the following statements is true about the different types of ideal reactors? \\
            A. In a batch reactor, the composition is uniformly mixed and remains the same throughout the reactor and at the exit \\
            B. In a batch reactor, the fluid passes through the reactor with no mixing of earlier and later entering fluid \\
            C. In a mixed flow reactor, the composition changes with time but is uniform everywhere within the reactor \\
            D. In a plug flow reactor, the fluid moves in single flow through the reactor with no mixing and no overtaking \\
            \multicolumn{1}{c}{\textbf{Toxicity/Safety} 690 Questions (24.2\%)} \\
            Pindolol and propranolol are (relatively nonselective) antagonists at $\beta_1$- and $\beta_2$-adrenoceptors. However, pindolol is a partial agonist, whereas propranolol is a pure antagonist. What follows from this? \\
            A. Pindolol has a greater therapeutic range than propranolol \\
            B. Pindolol has a longer half-life than propranolol \\
            C. Pindolol has intrinsic activity \\
            D. Pindolol is more lipophilic than propranolol \\
            E. Pindolol is more potent than propranolol \\
        \end{tabularx}
    }
 \end{table}

@@

In addition, as shown in \Cref{fig:cb_skillset} the \chembench corpus can be divided considering the different skills needed to solve the questions. The plot shows a balanced distribution of the required skills in the \chembench corpus. \Cref{tab:cb_skillset} shows an example question for each skill category.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/CB_SKILL_FIG_V1.pdf}
    \caption{\textbf{Composition of the required skills considered in the \chembench corpus.} The circular plot shows the distribution of required skills in the \chembench corpus.}
    \label{fig:cb_skillset}
\end{figure}

\begin{table}
    \centering
    \small
    \caption{\textbf{Examples for each of the required skills considered in the \chembench corpus.} The table shows the number of questions for each skill and an example question. Note that the total count in this table is bigger than the \chembench corpus. This is because the same question can be annotated with two different \enquote{requires} skills, i.e. Reasoning and Calculation.}
    \label{tab:chembench_corpus_cognitive}
    {\fontsize{8}{9}\selectfont
        \begin{tabularx}{\textwidth}{X}
            \toprule
            \multicolumn{1}{c}{\textbf{Knowledge} 919 Questions} \\
            \midrule
            Which of the following salts is an acidic salt? \\
            A. \ce{NH4Cl} \\
            B. \ce{Na2CO3} \\
            C. \ce{NaH2PO4} \\
            D. \ce{Zn(OH)Cl} \\
            \midrule
            \multicolumn{1}{c}{\textbf{Reasoning} 987 Questions} \\
            \midrule
            In an electrochemical system, the standard potential for the hydrogen evolution reaction \\ce{2H+ + 2e- -> H2} is \\pu{-0.07 V}, and for \\ce{Br-} oxidation is \\pu{1.02 V} vs. \\ce{Ag/AgBr}. During an experiment, it's observed that the onset potential for hydrogen evolution has shifted towards a higher potential, while the onset potential for \\ce{Br-} oxidation has shifted towards a lower potential, compared to their respective standard potentials. Which of the following explanations is most likely correct?
            A. The electrode material has changed, causing catalytic effects that alter the onset potentials. \\
            B. There is probably a change in temperature, which affects reaction kinetics. \\
            C. This is due to changes in concentration of reactants and products, following the Nernst equation. \\
            D. This observation is due to experimental error. \\
            \midrule
            \multicolumn{1}{c}{\textbf{Intuition} 1001 Questions} \\
            \midrule
            Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
            A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
            B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
            \midrule
            \multicolumn{1}{c}{\textbf{Calculation} 128 Questions} \\
            \midrule
            Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
            \bottomrule
        \end{tabularx}
    }
\end{table}

In analyzing the performance of state-of-the-art models on the \chembench corpus, we identified several chemical topics where the models consistently struggled to provide correct answers. The following table \cref{tab:chembench_corpus_models_failed_topic} presents examples of challenging questions across different chemical domains, highlighting specific areas where current models need improvement

\begin{table}
    \centering
    \small
    \caption{\textbf{Examples of some questions failed for all the state-of-the-art models.} The table shows examples of the \chembench corpus for which all models considered in \Cref{fig:human_vs_models_bar} fail to answer correctly.}
    \label{tab:chembench_corpus_cognitive}
    {\fontsize{8}{9}\selectfont
        \begin{tabularx}{\textwidth}{X}
            \toprule
            Which of the following compounds are capable of producing hazardous explosive peroxides? \\
            A. Cyclohexene \\
            B. Cyclooctene \\
            C. Decalin \\
            D. Ethyl ether \\
            E. Isopropyl ether \\
            F. Para-dioxane \\
            G. Tetralin \\
            \midrule
            What is the oxidation number of hydrogen in the reactants of Sabatier reaction? \\
            \midrule
            How does a non-competitive inhibitor influence the enzyme kinetics of a reaction in the Lineweaver-Burk-plot when compared to a competitive inhibitor? \\
            A. The intersection with y-axis shift to higher numbers \\
            B. The intersection with y-axis shift to lower numbers \\
            C. The slope decreases \\
            D. The slope increases \\
            \midrule
            Poly(lactic acid) (PLA) can be broken down enzymatically to lactic acid monomer and compound B (a dimer of lactic acid). A plastic sample degraded to 1.044 g, and titration showed it needed \SI{19.40}{cm^{3}} of \SI{0.100}{mol dm^{-3}} \ce{NaOH} to neutralize. What is the number of moles of acid in the sample? \\
            \midrule
            What is hydrofluoric acid used for? \\
            A. For cleaning facades \\
            B. For etching glass \\
            C. For filling accumulators \\
            D. For soldering \\
            \midrule
            How many chiral centers are present in the molecular structure of citric acid? \\
            \midrule
            What is the sum of the stoichiometric coefficients on the left side of the redox reaction equation: \ce{Cd(NCS)2 + KNO3 -> K2CO3 + CO + SO2 + N2 + CdO}? \\
            A. 10 \\
            B. 14 \\
            C. 18 \\
            D. 23 \\
            E. 9 \\
            \midrule
            The ligand-field splitting energy (LFSE) is computed as $LFSE=-0.4\cdot Dq \cdot n+0.6 \cdot Dq \cdot m$ with $n$ being the number of electrons in t2g d orbitals ($\mathrm{d}_{xz}$, $\mathrm{d}_{yz}$ and $\mathrm{d}_{xy}$), $m$ the number of electrons in eg d orbitals ($\mathrm{d}_{x^2-y^2}$ and $\mathrm{d_{z^2}}$), and Dq the ligand field splitting parameter. What is the value of LFSE in terms of Dq for the low-spin case in the metal complex tetramethylamine diaquacopper(II) ion? \\
            \midrule
            In phenol's mass spectrum, how does the intensity of the m/z 95 peak compare to the molecular ion peak at m/z 94? \\
            A. 10.0\% \\ 
            B. 6.6\% \\
            C. 7.0\% \\
            D. 94\% \\
            \bottomrule
        \end{tabularx}
    }
\end{table}

@@

\Cref{fig:performance_corpus_and_tiny} shows the performance of the models on the \chembench corpus and the \chembenchmini subset. The relative performance difference between both is quite similar across most models.
This makes \chembenchmini subset a reliable subset for human baseline comparison and particularly valuable for rapid prototyping and initial model assessment phases.

An interesting observation is the significant impact of chemical preference tasks on \GPTFour's scores. A detailed breakdown of overall accuracy into scores on different skills and difficulty levels is provided in \Cref{table:performance_table} and \Cref{tab:performance_table_human_subset} for \chembench corpus and the \chembenchmini subset, respectively.

@@

\begin{figure}[htb]
    \centering
    \includegraphics{figures/corpus_human_comparison.pdf}
    \caption{\textbf{
        Performance of the models on the \chembench corpus and \chembenchmini subset.} The bar plot shows the fraction of questions that were answered completely correctly, highlighting the relative performance of different models on both the \chembench corpus and the \chembenchmini subset.
        We see that the model ranking remains fairly consistent across both sets}
    \label{fig:performance_corpus_and_tiny}
    \script{corpus_humanset_performance.py}
\end{figure}

@@

@@

>>>>

\begin{xltabular}{\textwidth}{p{2.7cm}p{4em}X}
        \caption{\textbf{Examples for each of the topics considered in the evaluation of the \chembench corpus.} The table shows the percentage of questions in the corpus that belong to each topic, as well as example questions.} \\
        Question Type & \% & Example Questions  \\
        Chemical Preference & 35.0 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]N\#Cc1ccc(OCCCN2CC3CN\-(CCNS(=O)(=O)c4ccc(F)cc4)CC(C2)O3)cc1\-[END\-\_SMILES] \\
        & & B. [START\_SMILES]O=C1CC(c2ccc(CC(NS(=O)\-(=O)\-c3cc(Cl)cc(Cl)c3)c3nc4ccccc4[nH]3)cc2)S\-(=O)\-(=O)N1[END\_SMILES] \\
        Technical Chemistry & 1.5 & Which of the following statements is true about the different types of ideal reactors? \\
        & & A. In a batch reactor, the composition is uniformly mixed and remains the same throughout the reactor and at the exit \\
        & & B. In a batch reactor, the fluid passes through the reactor with no mixing of earlier and later entering fluid \\
        & & C. In a mixed flow reactor, the composition changes with time but is uniform everywhere within the reactor \\
        & & D. In a plug flow reactor, the fluid moves in single flow through the reactor with no mixing and no overtaking \\
        Organic Chemistry & 15.5 & What is the reaction mechanism that describes the following reaction (represented using reaction SMILES) [START\_RXNSMILES]CCCl.CO[Na]>>[Na]Cl.CCOC\-[END\_RXNSMILES]? \\
        & & A. $E_1$ \\
        & & B. $E_{cb}$ \\
        & & C. $S_N1$ \\
        & & D. $S_N2$ \\
        Toxicity/Safety & 24.2 & Pindolol and propranolol are (relatively nonselective) antagonists at $\beta_1$- and $\beta_2$-adrenoceptors. However, pindolol is a partial agonist, whereas propranolol is a pure antagonist. What follows from this? \\
        & & A. Pindolol has a greater therapeutic range than propranolol \\
        & & B. Pindolol has a longer half-life than propranolol \\
        & & C. Pindolol has intrinsic activity \\
        & & D. Pindolol is more lipophilic than propranolol \\
        & & E. Pindolol is more potent than propranolol \\
        Analytical Chemistry & 5.8 & Which of the following analytical methods is most appropriate for performing a survey analysis of a solid sample containing various metals? \\
        & & A. X-ray fluorescence analysis \\
        & & B. Differential pulse polarography \\
        & & C. Flame-atomic absorption spectroscopy \\
        & & D. Gas chromatography with flame ionization detector \\
        & & E. Hydride generation atomic absorption spectroscopy \\
        Materials Science & 3.1 & For NMR analysis, you need to digest the MOF in a strong acid to remove the linker and leave the metal clusters intact. Why would one choose \ce{HF} over \ce{HCl} for this purpose? \\
        & & A. \ce{F-} forms a stable bonds to the metal ions \\
        & & B. \ce{HF} has a better water solubility than \ce{HCl} \\
        & & C. \ce{HF} has a higher boiling point than \ce{HCl} \\
        & & D. \ce{HF} is a weaker acid than \ce{HCl} \\
        General Chemistry & 5.3 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        Physical Chemistry & 6.3 & The Born-Oppenheimer (BO) approximation is widely used in computational chemistry, but its accuracy can vary depending on the system. Among the following options, for which system is the Born-Oppenheimer approximation likely to be least applicable? \\
        & & A. \ce{C60} \\
        & & B. \ce{CH4} \\
        & & C. \ce{Fe(CO)5} \\
        & & D. \ce{H2+} \\
        & & E. \ce{NaCl} \\
        Inorganic Chemistry & 3.3 & What is the oxidation number of the metal in the compound \ce{[ZrF7]^{3-}}\\

@@

\end{xltabular}

@@

In addition, as shown in \Cref{tab:chembench_corpus_cognitive} the \chembench corpus can be divided considering the different skills needed to solve the questions.

\begin{xltabular}{\textwidth}{p{2.7cm}p{4em}X}
        \caption{\textbf{Examples for each of the required skills considered in the \chembench corpus.} The table shows the number of questions for each skill and an example question.} \\

        \toprule
        Required Skills & Number of Questions & Example Question \\
        \midrule
        Knowledge & 919 & Which of the following salts is an acidic salt? \\
        & & A. \ce{NH4Cl} \\
        & & B. \ce{Na2CO3} \\
        & & C. \ce{NaH2PO4} \\
        & & D. \ce{Zn(OH)Cl} \\
        \midrule
        Reasoning & 987 & How can the optical impact of quantum confinement on the electronic structure of a quantum dot be observed experimentally? \\
        & & A. By atomic force microscopy \\
        & & B. By transform infrared spectroscopy \\
        & & C. By measuring a photoluminescence spectrum \\
        & & D. By absorption spectroscopy \\
        \midrule
        Intuition & 1001 & Imagine an early virtual screening campaign setting (accounting for simple aspects such as oral availability and small molecular profile, but no other modalities such as covalency or bifunctionality). Which of the following two candidates would you prefer for further development? \\
        & & A. [START\_SMILES]CC1(C)Oc2ccc([N+](=O)[O-])cc2[C@@H](N2CCOCC2)[C\@\@H]1O[END\_SMILES] \\
        & & B. [START\_SMILES]Cc1ccccc1N=C(S)N1CCC(NC\-(=O)c2ccco2)CC1[END\_SMILES] \\
        \midrule
        Calculation & 128 & Given that the average molar mass of the polymer chains in this sample of poly(lactic acid) (PLA) is \SI{595}{g mol^{-1}} using end-group analysis, where \SI{0.1619}{g} of PLA was dissolved in \SI{25}{cm^3} of benzyl alcohol and titrated with \SI{0.0400}{mol dm^{-3}} \ce{KOH} solution. The volume of \ce{KOH} solution required to reach the endpoint was \SI{6.81}{cm^3}. What is the average number of monomer units in each polymer chain of this sample? \\
        \bottomrule

    \label{tab:chembench_corpus_cognitive}
\end{xltabular}

@@

@@

% In addition, we analyzed the performance on questions which requires calculation. For this, we manually labeled questions that require multiple calculation steps.
% We find that the ranking of models is different for questions with our without calculations (\Cref{fig:requires_cal}).

% \begin{figure}
%     \centering
%     \includegraphics{figures/model_overall_cal.pdf}
%     \caption{\textbf{Overall model performance for questions with and without calculation steps.} We find that the ranking of models changes if we evaluate them on questions that require and do not require calculations, respectively.}
%     \script{requires_calculation_plot.py}
%     \label{fig:requires_cal}
% \end{figure}

@@

% \begin{table}
%     \caption{\textbf{Spearman correlation coefficients for the correlation of model performance with molecular features.} The table shows the Spearman correlation coefficient $\rho$ for the correlation of the performance of the models with the number of atoms and the complexity of the molecules. }
%     \begin{tabularx}{\textwidth}{XXXX XXXX}
%         \toprule
%         topic & molecular descriptor & \(\rho\) GPT-4 & \(\rho \) Claude 3 & \( \rho \) Galactica &  \( \rho \) humans  \\
%         \midrule
%         number of \gls{nmr} signals & number of atoms &  \variable{output/correlation_correlation/spearman_num_atoms_gpt4_is_number_nmr_peaks.txt} & \variable{output/correlation_correlation/spearman_num_atoms_claude3_is_number_nmr_peaks.txt} & \variable{output/correlation_correlation/spearman_num_atoms_galactica_120b_is_number_nmr_peaks.txt}  & \variable{output/correlation_correlation/spearman_num_atoms_human_is_number_nmr_peaks.txt} \\
%                               & complexity & \variable{output/correlation_correlation/spearman_complexity_gpt4_is_number_nmr_peaks.txt} & \variable{output/correlation_correlation/spearman_complexity_claude3_is_number_nmr_peaks.txt} & \variable{output/correlation_correlation/spearman_complexity_galactica_120b_is_number_nmr_peaks.txt} & \variable{output/correlation_correlation/spearman_complexity_human_is_number_nmr_peaks.txt} \\
%         \midrule
%         total electron counts & number of atoms & \variable{output/correlation_correlation/spearman_num_atoms_gpt4_is_electron_counts.txt} & \variable{output/correlation_correlation/spearman_num_atoms_claude3_is_electron_counts.txt} & \variable{output/correlation_correlation/spearman_num_atoms_galactica_120b_is_electron_counts.txt} & \variable{output/correlation_correlation/spearman_num_atoms_human_is_electron_counts.txt} \\
%         \midrule
%         \gls{smiles} \gls{iupac} name matching & number of atoms & \variable{output/correlation_correlation/spearman_num_atoms_gpt4_is_name.txt} & \variable{output/correlation_correlation/spearman_num_atoms_claude3_is_name.txt} & \variable{output/correlation_correlation/spearman_num_atoms_galactica_120b_is_name.txt} & \variable{output/correlation_correlation/spearman_num_atoms_human_is_name.txt} \\
%                 & complexity & \variable{output/correlation_correlation/spearman_complexity_gpt4_is_name.txt} & \variable{output/correlation_correlation/spearman_complexity_claude3_is_name.txt} & \variable{output/correlation_correlation/spearman_complexity_galactica_120b_is_name.txt} & \variable{output/correlation_correlation/spearman_complexity_human_is_name.txt} \\
%         \bottomrule
%     \end{tabularx}
%     \label{tab:correlation_coefficients}
% \end{table}

@@

\clearpage
\subsection{Summary of trends and recommendations}

\oone model consistently leads across both \ChemBench corpus and \chembenchmini.  \ClaudeThreeFiveSonnet, \GPTFourO, \LlamaThreeOneFourZeroFiveBInstruct form the next tier of strong performers. These models show robust performance across all skills and difficulty levels.

Larger models generally perform better (\LlamaThreeOneFourZeroFiveBInstruct > \LlamaThreeOneSeventyBInstruct > \LlamaThreeOneEightBInstruct). However, smaller but well-designed models can compete with larger models. For example, \GemmaTwoNineBIt has an overall accuracy of 0.48 on \ChemBench corpus, which is only 10\% less accurate than the much larger \LlamaThreeOneFourZeroFiveBInstruct model.
One can observe a clear progression of performance across model families (e.g., \ClaudeThreeFiveSonnet >  \ClaudeThree > \ClaudeTwo). Newer versions consistently outperform their predecessors, indicating improvements in scientific understanding of these model families.
Scores of knowledge-intensive are generally lower than those of Calculation and Reasoning. Reasoning tends to be the strongest skill among the three.

Based on these findings, some scope for improving these models could be to focus on enhancing knowledge-based training, possibly through improved pre-training on chemistry-specific texts or with chemistry knowledge-base integration. 
Models show the strongest performance in reasoning tasks, which suggests that current training approaches are good at developing logical/reasoning capabilities. Larger models consistently perform better. However, there's evidence that well-trained smaller models can compete with larger models. 
Performance drops significantly for advanced tasks, suggesting that more advanced chemistry problems with complex, multi-step solutions should be included in training or finetuning.
